+ solar location names
+ NCBI data
  + Computing Column Widths
  + Finding the tables
  + Cleaning the text
+ Reading Web logs. (Data/Weblogs/strategy.Rmd)


## Case Study - Solar station location names

Consider the names of the files in the ZIP files from assignment 1.

We want more human-readable text for the locations for the .stat files.

First, we have to find the .stat files i.e., end in .stat 

```r
fnames = grep("\\.stat$", list.files(DataDir), value = TRUE)
fnames = list.files(DataDir, pattern = "\\.stat$")
```

Or
```r
w = tools::file_ext(list.files(DataDir)) == "stat"
fnames = list.files(DataDir)[w]
```
Look at the implementation of  tools::file_ext - regexpr.


So we have 
```r
fnames
```
```
[1] "USA_CA_Fairfield-San.Francisco.Bay.Reserve.998011_TMYx.2007-2021.stat"
[2] "USA_CA_Marin.County.AP-Gnoss.Field.720406_TMYx.2007-2021.stat"        
[3] "USA_CA_Napa.County.AP.724955_TMYx.2007-2021.stat"                     
[4] "USA_CA_Point.Reyes.Lighthouse.724959_TMYx.2007-2021.stat"             
[5] "USA_CA_UC-Davis-University.AP.720576_TMYx.2007-2021.stat"             
```

We want 
+ remove the USA_CA_
+ remove the TMYx.2007-2021.stat
+ remove the 6 digit number
+ remove the AP if it precedes the 6 digit number


```r
ids = gsub("USA_CA_|TMYx.2007-2021.stat|(AP\\.)?[0-9]{6}_", "", fnames)
```

Don't want the . at the end.
+ either remove now
  ids = gsub("\\.$", "", ids)
+ or in the original gsub()

```r
ids = gsub("USA_CA_|TMYx.2007-2021.stat|\\.(AP\\.)?[0-9]{6}_", "", fnames)
```

Replace . with space
```r
gsub("\\.", " ", ids)
```
```
[1] "Fairfield-San Francisco Bay Reserve" "Marin County AP-Gnoss Field"        
[3] "Napa County"                         "Point Reyes Lighthouse"             
[5] "UC-Davis-University"                
```

----------

Let's go back to the NCBI data.

We'll see how to use regular expressions in several different ways
for several different steps in reading the two tables

+ finding the tables
+ cleaning the values before we call read.fwf()
   + specifically, removing the % in the cover column so that read.fwf() will convert the values to
     a numeric vector and we don't have to post process the column.
+ finding the starting location of each column so we can compute the column widths


## Computing the Column Widths

How can we  find the column widths
I showed a technique in funs2.R using strsplit() and matching characters.

The idea was
+ Take the header line start with Description
+ find the positions in the line that
   + start with a capital letter
   + lower case letter but are preceded by a space, i.e., " c"

Let's get the column header line
```r
h = txt2[1]
```

We want to find positions in the string `h` corresponding to patterns we want to match.
So we use gregexpr(), not grep() which finds indices in vectors, not within strings.

We'll build up our regular expression/pattern to match iteratively, testing it each step of the way.

We start by finding the capital letters.
We can use a character class for that `[A-Z]`

```r
pos = gregexpr("[A-Z]", h, perl = TRUE)[[1]]
length(pos)
pos
```

All matches have match.length 1


What about adding lower case letters
```r
pos = gregexpr("[A-Za-z]", h, perl = TRUE)[[1]]
length(pos)
```
We got way too many, of course!
We matched every lower case letter in each of the column name text.


What about just 'c'
```r
pos = gregexpr("[A-Zc]", h, perl = TRUE)[[1]]
length(pos)
```
+ Matches c in cover - good
+ Matches c in Accession - twice - bad


We need to be more specific

It is EITHER
+ a capital letter
+ OR
+ a space followed by a lower case letter

```r
pos = gregexpr("([A-Z]| c)", h, perl = TRUE)[[1]]
length(pos)
```


Looks good, but this tells us the start of the cover column is at 122, but it should be 123.
The match starts at the space, not at c.

The match has length 2, while the others have length 1.

We could adjust by the length of the match.
+ If it is length 1, add nothing.
+ If it is length 2, add 1

```r
pos2 = pos + attr(pos, "match.length") - 1L
```
Check this gives us what we expect
```r
substring(h, pos2, pos2)
```


We can be a little fancier if we didn't count the space preceding the c as part of the match,
but just require it to be there.

If we could match a lower case letter and then look-behind/back at the previous character(s) for a pattern, that would do it.
We'd match the c and not the space as the start of our match.

We can specify a lookbehind via
```r
(?<=pattern)
```
Note, we need perl = TRUE in the call to grep/gsub/gregexpr to use lookbehinds.

So our full pattern is 
```r
(?<= )c
    ^
  space
```

Putting this all together, we have
```
pos3 = gregexpr("[A-Z]|(?<= )[a-z]", h, perl = TRUE)[[1]]
```

Now we are ready to compute the column widths and call read.fwf()
```
widths = diff(c(pos3, nchar(h)))

d3 = read.fwf(textConnection(txt2[-1]), widths)
```

---



## Finding the Tables

```{r}
nc = readLines("~/sta141b/Data/NCBI/NCBIQuery.txt")
```

Let's find the lines that contain the word Description
```{r}
idx = grep("Description", nc)
txt = grep("Description", nc, value = TRUE)
w = grepl("Description", nc)
```
So grep() can give the positions in the vector that had a match, or return the actual matching
values.

And grepl() returns a logical vector of TRUE and FALSE values indicating which matched.



Next we'll match the lines that start with Description or that are blank, i.e., the start is
followed by the end of the line.

```{r}
w = grepl("(^Description|^$)", nc)
```
This is equivalent to 
```{r}
w2 = grepl("^Description", nc) | grepl("^$", nc)
table(w, w2)
```

Now, here comes a useful trick.
We have a logical vector in which TRUE values mark the start of a "section"
or the start of the next section.

A line starting with Description identifies the start of a table.
A blank line identifies the start of the next section after a table, or just   a blank line.

We can compute the cumulative sum of the TRUE and FALSE values.
Whenever there is a TRUE, the cumsum increases by 1. For the FALSE values, the cumsum stays the
same.
So the cumsum changes for each point of interest.
So we will end up with something like
```{r}
0,0,0, 1, 1, 1, 1, ..., 1, 2, 2, 2, ....
```
These identify which group each line is in and we can split by group identifier with
```{r}
g = split(nc, cumsum(w))
```

There are lot of these, but we only want the tables, i.e., the one whose first
element is the line that starts with (or contains) Description

```{r}
isTbl = sapply(g, function(x) grepl("Description", x[1]))
table(isTbl)
tbls = g[isTbl]
```

Let's look at the first table
```{r}
tbls[[1]]
```


## Cleaning the Text

Let's clean the numeric values *before* passing to read.fwf().
Specifically, remove the % after a numeric value so read.fwf() can convert the column to numbers if possible.

```{r}
txt = tbls[[1]]
```

We'll use gsub() to remove %.
We could do it everywhere, but we want to replace it only when
it is immediately after digits/numbers.
We match the pattern and then specify what the new content should be to replace the specific 
match.
We want to match 1 or more digits then followed by a %.

We could use
"(0|1|2|3|4|5|6|7|8|9)+%"
which is 
+ any of these 10 digits and then 
+ 1 or more of these 
+ followed by the literal %

We can use a character class to avoid the lengthy "(0|1|2|3|4|5|6|7|8|9)"
Specifically, we can use  `[0-9]` or `[[:digits:]]` or

```
rx = "[0-9]+%"
```

We don't need the () as the quantifier + applies to the preceding subpattern `[0-9]`.
However, we put () around the `[0-9]+` to "capture" the matched pattern.
We do this so we can refer to it in the pattern we use to specify how to transform the match, i.e.,

```
rx = "([0-9]+)%"
```

The % is not in the () because we want to reference only the number part of each match.

We can reference the number part of each match with \\1 in the second argument of 

```{r}
txt2 = gsub(rx, "\\1", txt)
```

If these contained decimal values, we would allow a .

txt2 = gsub("([0-9.]+)%", "\\1", txt)

Try it on some sample data.
gsub("([0-9.]+)%", "\\1", c("100%", "98.756%", "50"))




Having cleaned the text, we can pass it to read.fwf()
But we need the column widths
```
read.fwf(textConnection(txt2), widths = ???)
```



